\documentclass[letterpaper, 10 pt, conference]{ieeeconf}  % Comment this line out if you need a4paper

%\documentclass[a4paper, 10pt, conference]{ieeeconf}      % Use this line for a4 paper

\IEEEoverridecommandlockouts                              % This command is only needed if 
                                                          % you want to use the \thanks command

\overrideIEEEmargins                                      % Needed to meet printer requirements.

% See the \addtolength command later in the file to balance the column lengths
% on the last page of the document

% The following packages can be found on http:\\www.ctan.org
\usepackage{times}
\usepackage{cite}
\usepackage{graphics} % for pdf, bitmapped graphics files
\usepackage{graphicx}
\usepackage{epsfig} % for postscript graphics files
\usepackage{mathptmx} % assumes new font selection scheme installed
%\usepackage{times} % assumes new font selection scheme installed
\usepackage{amsmath} % assumes amsmath package installed
\usepackage{amsfonts}
\usepackage{amssymb}  % assumes amsmath package installed
\usepackage{bm}
\usepackage{soul}
% \usepackage{todonotes}
\usepackage{url}
\usepackage{multicol}
\usepackage{multirow}
\usepackage[bookmarks=true]{hyperref}
\usepackage{verbatim}
\usepackage{color}
\usepackage{tabularx}
\usepackage{booktabs}
\usepackage{listings}
\usepackage{siunitx}
\usepackage[utf8]{inputenc}
\usepackage{pifont}
\usepackage{csquotes}
\usepackage{mathtools}
\usepackage{breqn}
\usepackage{cancel}
\usepackage{xcolor}
\usepackage{optidef}
\usepackage{cleveref}
\usepackage{setspace}

\newcommand{\cmark}{\ding{51}}%
\newcommand{\xmark}{\ding{55}}%

%\makeatletter
%\def\endthebibliography{%
%  \def\@noitemerr{\@latex@warning{Empty `thebibliography' environment}}%
%  \endlist
%}
%\makeatother

\begin{document}
	
\linespread{0.98}
\setlength{\textfloatsep}{1pt}  % Replace X with the desired length in points
\setlength{\intextsep}{1pt}
\setlength{\abovecaptionskip}{1pt}

\title{\LARGE \bf
Optimal Design of Agile Jumping Maneuvers for a Single Leg System
% heavy-duty 
% Impact-Aware Planning and Control of Agile Jumping Maneuvers: from the Thrust Phase to Landing
%Designing Aerial Maneuvers for Agile Legged Systems via Trajectory Optimization: from take-off to landing
%Design and Control of Jumping Maneuvers via Trajectory Optimization: from the Thrust Phase to Landing
}

\author{$^{1\,-\,2}$, Arturo Laurenzi$^{1}$ $^1$
% , Francesco Ruscelli1^{1}, Luca Rossini$^{1}
\thanks{  Please send correspondence to \tt\scriptsize \{andrea.patrizi\}@iit.it}}
\author{%
  \authorblockN{%
    Andrea Patrizi\authorrefmark{1}\authorrefmark{2},
    Francesco Ruscelli\authorrefmark{1}\authorrefmark{3}, 
    Arturo Laurenzi\authorrefmark{1}\authorrefmark{3},
    and Nikos G. Tsagarakis\authorrefmark{1}
  }
%  \authorblockA{\authorrefmark{1} Humanoid and Human-Centred Mechatronics (HHCM), Istituto Italiano di Tecnologia (IIT), Via San Quirico 19d, 16163 Genova}%
%  \authorblockA{\authorrefmark{2} }
}

% footnotes
\maketitle
\begingroup\renewcommand\thefootnote{\authorrefmark{2}}
\footnotetext{Department of Informatics, Bioengineering, Robotics and Systems Engineering, Università di Genova, Via All'Opera Pia 13, 16145 Genova.}
\endgroup
\begingroup\renewcommand\thefootnote{\authorrefmark{1}}
\footnotetext{Humanoids and Human-Centred Mechatronics (HHCM), Istituto Italiano di Tecnologia (IIT), Via San Quirico 19d, 16163 Genova.}
\endgroup
\begingroup\renewcommand\thefootnote{\authorrefmark{3}}
\footnotetext{The authors contributed equally to the manuscript.}
\endgroup

\setlength{\textfloatsep}{12.0pt plus 8.0pt minus .0pt}

%\begingroup\renewcommand\thefootnote{}
%\footnotetext{\{francesco.ruscelli, arturo.laurenzi,
%enrico.mingo, nikos.tsagarakis\}@iit.it}
%\endgroup
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
In this work we showcase a comprehensive study, developed around our heavy-duty eletrically-actuated quadruped leg prototype, focusing on the task of jumping. Our study employs a two-stage pipeline for the design and execution of agile and efficient jumping maneuvers.
We combine offline non-linear whole-body Trajectory Optimization (TO)
%, which considers the robot's multi-body dynamics and calibrated actuators models
to formulate and solve two distinct non-linear problems, each tackling different phases and performance aspects of the jumping maneuver. 
The first one jointly addresses the thrust and flight phases and aims at maximizing the achieved jump height, while accounting for several hardware limitations.
The second problem deals with the impact and braking phases 
%and optimizes for a landing configuration and joint-space impedances,
%with the objective of 
minimizing the kinetic energy dissipated at the impact, while maximizing the energy regenerated by the actuators during braking. We 
%explore the driving reasons behind our formulations and approach and 
validate the accuracy of the employed models and the results of the optimizations both in simulation and on the real leg hardware, showing the effectiveness of our approach in maximizing hardware performance and the benefits of exploiting combined impact-energy considerations during the execution of jumping maneuvers.
\end{abstract}

\IEEEpeerreviewmaketitle


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\input{introduction}

\input{problem_definition}

% \input{contributions}

\input{formulation}

% \input{simulation_results}

\input{experimental_results}

\input{conclusions}

\input{acknowledgements}

% \input{appendix}
\bibliographystyle{ieeetr}\vspace{-0.2cm}
\bibliography{bibliography/cartesian_imp,bibliography/impact_dyn,bibliography/rigid_body_algo,bibliography/joint_friction_estimation,bibliography/force_estimation,bibliography/misc,bibliography/biomechanical_refs,bibliography/regenerative_braking,bibliography/foc_cntrl_bldc,bibliography/agile_robots_design_and_cntrl,bibliography/to,bibliography/xbot,bibliography/urls}

\end{document}
